### 2023-10-03 writing/new-paper/overview.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

There is a subtlety here: The layout of a subexpression inside of a
function definition might not be determined by the layouts that the
function definition is instantiated at. In this case, we must
explicitly specify a layout. This is done using the \verb|instantiate|
and \verb|lower| constructs. The former is for function calls, while
the latter is for all other types of values. These operations are
performed with the \verb|instantiate| and \verb|lower| language
constructs, respectively. For example, \verb|instantiate [A] B f x|
calls $f$ with the argument $x$, where the layout $A$ is used for the
argument and the layout $B$ is used for the result of the function
application. This is also written $\instantiateS{A,B}{f}(x)$. The
latter is primarily used in \autoref{sec:semantics}.

The code generator is instructed to generate a \SuSLik{} specification
for a certain function at a certain instantiation by using a
\generate{} directive. For example, if there is a function definition
with the type signature \lstinline[language=Pika]{mapAdd1 : List -> List}, a line reading
\lstinline[language=Pika]{$\%$generate mapAdd1 [Sll] Sll} would instruct the
\Pika{} compiler to generate the \SuSLik{} inductive predicate
corresponding to \verb|mapAdd1| instantiated to the \verb|Sll| layout
for both its argument and its result.

@@

We demonstrate the characteristic usages of \Pika{} on a series of
examples.
%
In these examples, we will often make use of the \verb|List| algebraic
data type and its \verb|Sll| layout from~\autoref{fig:List-def}. A
simple example of \Pika{} code that illustrates algebraic data types and
layouts is a function which creates a singleton list out of the given
integer argument:

@@

\noindent
A slightly more complicated example comes from trying to write a
functional-style \verb|map| function directly in \SuSLik. Consider a
function which adds 1 to each integer in a list of integers.
Considering the list implementation to be a singly-linked list with a
fixed layout, one way to express this in \SuSLik{} is shown in \autoref{fig:sllx}

@@

\noindent
Note that inductive predicates are used for \textit{two} different purposes: the \verb|sll| inductive
predicate describes a singly-linked list data structure, while the \verb|mapAdd1| inductive predicate 
describes how the input list relates to the output list. Both are used in the specification of
\verb|mapAdd1_fn|: \verb|sll| in the precondition and \verb|mapAdd1| in the postcondition.

Using the \verb|mapAdd1| inductive predicate gives us two advantages over attempting to
put the SSL propositions directly into the postcondition of \verb|mapAdd1_fn|:
%
\begin{enumerate}
  \item We are able to express a conditional on the shape of the list. This is much like pattern
    matching in a language with algebraic data types, but we are examining the pointer involved directly.
  \item We are able to express \textit{recursion} part of the postcondition: the \verb|mapAdd1| inductive
    predicate refers to itself in the \verb|not (x == 0)| branch.
\end{enumerate}
%
These two features are both reminiscent of features common in
functional programming languages: pattern matching and recursion.
However, there are still some significant differences:

@@

Say we want to write the functional program that corresponds to this specification. One way to do this in a Haskell-like language
is the following, using the \verb|List| type from \autoref{fig:List-def}.

@@

\noindent
The only missing information is the memory representation of the \verb|List| data structure. We do not want the
\verb|mapAdd1_fn| implementation to deal with this directly, however. We want to separate the more abstract notions
of pattern matching and constructors from the concrete memory layout that the data structure has.

To accomplish this, we now extend the code with the definition of
\verb|Sll| from \autoref{fig:List-def}. \verb|Sll| is a
\textit{layout} for the algebraic data type \verb|List|.
%
Now we have all of the information of the original specification but
rearranged so that the low-level memory layout is separated from the
rest of the code. This separation brings us to an important
observation about the language, manifested throughout these examples:
none of the function definitions need to \emph{directly} perform any
pointer manipulations. This is relegated entirely to the reusable
layout definitions for the ADTs. The examples are written entirely as
recursive functions that pattern match on, and construct, ADTs.

All that is left is to connect these two parts: the layouts and the
function definitions. We instruct a \SuSLik{} specification generator to
generate a \SuSLik{} specification from the \verb|mapAdd1_fn| function
using the \verb|Sll| layout:
%
\begin{lstlisting}[language=Pika]
%generate mapAdd1_fn [Sll] Sll
\end{lstlisting}
%
The \verb|[Sll]| part of the directive tells the generator which
layouts are used for the arguments. In this case, the function only
has one argument and the \verb|Sll| layout is used. The \verb|Sll| at
the end specifies the layout for the result.

@@

We can generalise our \verb|mapAdd1| to map arbitrary \verb|Int|
functions over a list and then redefine \verb|mapAdd1| using the new
\verb|map|.

@@

This example makes use of \lstinline{instantiate} in two places. In the first case where we have the call
\lstinline{instantiate [Int] Int f x}, the builtin \verb|Int| layout is used for both the input
and output. In this special case, the \verb|Int| layout shares a name with the \verb|Int| type
that it represents. This is necessary since \lstinline{instantiate} is used for all non-recursive calls that are not constructor applications.

In the second use, \lstinline{instantiate [Int -> Int, Sll] Sll map add1 xs}, we specify that
the second argument uses the \verb|Sll| layout for the \verb|List| type from \autoref{fig:List-def}. We also give \verb|Sll| as
the layout for the result of the call. Note that it is not necessary to use \verb|instantiate| for the recursive call to \verb|map|. This is because the appropriate layout
is inferred for recursive calls.

It might be surprising that \verb|instantiate| is required in the body of \verb|mapAdd1| since the type signature
of \verb|mapAdd1| suggests that it is layout polymorphic and yet we must pick a specific \verb|List| layout when
we use \verb|instantiate| to call \verb|map|. This is because, in general, a call inside the body of some function \verb|fn| might use
any layout, even layouts that have no relation to the layouts that \verb|fn| is instantiated to.

It is possible to do inference of some layouts, for example in
\verb|mapAdd1| we would usually want to use the same layout as the
argument layout, but we leave this for future work. Another approach
is to introduce type variables that correspond to layouts, as done in
the series of works on the \tname{Dargent}{}
tool~\cite{chen:2023:dargent}.
%
We leave this approach for future work as well.

@@

Two more features that are common in functional languages are \verb|let| bindings and
\verb|if-then-else| expressions. Both of these have straightforward translations into
\SuSLik. A \verb|let| binding corresponds to introducing a new variable with an equality
constraint corresponding to the equality in the \verb|let| binding in the pure part of a \SuSLik{} assertion.

The \verb|if-then-else| construct corresponds to \SuSLik's C-like ternary operator. We can use these features together to implement a \verb|maximum| function:

@@

\noindent
The \verb|let| binding is required here. This is because the recursive predicate
application must use an additional variable, beyond the argument and output
parameters, to store the intermediate result \verb|i| and this is introduced in
\Pika{} using a \verb|let|. It is possible to extend the translator to
automatically introduce these \verb|let| bindings, but we relegate this to future work.

Furthermore, since the \verb|let| is required and the variable introduced by the \verb|let| is used
in the \verb|if| condition, we cannot replace the \verb|if| with a guard. Guards can only occur
at the top-level of a function definition, so it could not occur after a \verb|let|.

@@

A fold is a common kind of operation on a data structure in functional programming, where a binary function
is applied the elements of a data structure to create a summary value. For example, if the binary function
is the addition function, it will give the sum of all the elements of the data structure. The
classic example of such a fold is a fold on a list. In this example, we will write a
right fold over a \verb|List|.

We also demonstrate \verb|Ptr| types and the \verb|addr| builtin operation. This is somewhat similar to an \verb|@| pattern in Haskell,
but it is a low-level construct, indictating how pointers should be used, rather than a high-level construct.
Given a base type $\alpha$ (such as \verb|Int|), \verb|Ptr Int| is a type. This type is
passed around differently in the generated code. In particular, it is passed by reference.
\verb|addr| is the corresponding value-level introduction operation.
\begin{lstlisting}[language=Pika]
%generate fold_List [Int, Sll] (Ptr Int)

@@

\noindent
The compiler produces the following \SuSLik{} specification for \verb|fold_List|:

\begin{lstlisting}[language=SynLang]
predicate fold_List(int i, loc p_x ,loc __r) {
| (p_x == 0) => { __r :-> i }
| (not (p_x == 0)) => {
  p_x :-> x ** (p_x+1) :-> xs ** [p_x,2] **
  func f(p_x, __p_2, __r) **
  fold_List(i, xs, __p_2) **
  temploc __p_2 }
}
\end{lstlisting}
%
In the generated code, \verb|f| is applied to \verb|p_x| rather than
\verb|x| as a result of using \verb|Ptr| and \verb|addr|.

>>>>

There is a subtlety here: The layout of a subexpression inside of a
function definition might not be determined by the layouts that the
function definition is instantiated at. In this case, we must
explicitly specify a layout. This is done using the \lstinline[language=Pika]{instantiate}
and \lstinline[language=Pika]{lower} constructs. The former is for function calls, while
the latter is for all other types of values. These operations are
performed with the \lstinline[language=Pika]{instantiate} and \lstinline[language=Pika]{lower} language
constructs, respectively. For example, \lstinline[language=Pika]|instantiate [A] B f x|
calls $f$ with the argument $x$, where the layout $A$ is used for the
argument and the layout $B$ is used for the result of the function
application. This is also written $\instantiateS{A,B}{f}(x)$. The
latter is primarily used in \autoref{sec:semantics}.

The code generator is instructed to generate a \SuSLik{} specification
for a certain function at a certain instantiation by using a
\generate{} directive. For example, if there is a function definition
with the type signature \lstinline[language=Pika]{mapAdd1 : List -> List}, a line reading
\lstinline[language=Pika]{$\%$generate mapAdd1 [Sll] Sll} would instruct the
\Pika{} compiler to generate the \SuSLik{} inductive predicate
corresponding to \lstinline[language=Pika]{mapAdd1} instantiated to the \lstinline[language=Pika]{Sll} layout
for both its argument and its result.

@@

We demonstrate the characteristic usages of \Pika{} on a series of
examples.
%
In these examples, we will often make use of the \lstinline[language=Pika]{List} algebraic
data type and its \lstinline[language=Pika]{Sll} layout from~\autoref{fig:List-def}. A
simple example of \Pika{} code that illustrates algebraic data types and
layouts is a function which creates a singleton list out of the given
integer argument:

@@

\noindent
A slightly more complicated example comes from trying to write a
functional-style \lstinline[language=Pika]{map} function directly in \SuSLik. Consider a
function which adds 1 to each integer in a list of integers.
Considering the list implementation to be a singly-linked list with a
fixed layout, one way to express this in \SuSLik{} is shown in \autoref{fig:sllx}

@@

\noindent
Note that inductive predicates are used for \textit{two} different purposes: the \lstinline[language=Pika]{sll} inductive
predicate describes a singly-linked list data structure, while the \lstinline[language=Pika]{mapAdd1} inductive predicate 
describes how the input list relates to the output list. Both are used in the specification of
\lstinline[language=Pika]{mapAdd1_fn}: \lstinline[language=Pika]{sll} in the precondition and \lstinline[language=Pika]{mapAdd1} in the postcondition.

Using the \lstinline[language=Pika]{mapAdd1} inductive predicate gives us two advantages over attempting to
put the SSL propositions directly into the postcondition of \lstinline[language=Pika]{mapAdd1_fn}:
%
\begin{enumerate}
  \item We are able to express a conditional on the shape of the list. This is much like pattern
    matching in a language with algebraic data types, but we are examining the pointer involved directly.
  \item We are able to express \textit{recursion} part of the postcondition: the \lstinline[language=Pika]{mapAdd1} inductive
    predicate refers to itself in the \lstinline[language=Pika]{not (x == 0)} branch.
\end{enumerate}
%
These two features are both reminiscent of features common in
functional programming languages: pattern matching and recursion.
However, there are still some significant differences:

@@

Say we want to write the functional program that corresponds to this specification. One way to do this in a Haskell-like language
is the following, using the \lstinline[language=Pika]{List} type from \autoref{fig:List-def}.

@@

\noindent
The only missing information is the memory representation of the \lstinline[language=Pika]{List} data structure. We do not want the
\lstinline[language=Pika]{mapAdd1_fn} implementation to deal with this directly, however. We want to separate the more abstract notions
of pattern matching and constructors from the concrete memory layout that the data structure has.

To accomplish this, we now extend the code with the definition of
\lstinline[language=Pika]{Sll} from \autoref{fig:List-def}. \lstinline[language=Pika]{Sll} is a
\textit{layout} for the algebraic data type \lstinline[language=Pika]{List}.
%
Now we have all of the information of the original specification but
rearranged so that the low-level memory layout is separated from the
rest of the code. This separation brings us to an important
observation about the language, manifested throughout these examples:
none of the function definitions need to \emph{directly} perform any
pointer manipulations. This is relegated entirely to the reusable
layout definitions for the ADTs. The examples are written entirely as
recursive functions that pattern match on, and construct, ADTs.

All that is left is to connect these two parts: the layouts and the
function definitions. We instruct a \SuSLik{} specification generator to
generate a \SuSLik{} specification from the \lstinline[language=Pika]{mapAdd1_fn} function
using the \lstinline[language=Pika]{Sll} layout:
%
\begin{lstlisting}[language=Pika]
%generate mapAdd1_fn [Sll] Sll
\end{lstlisting}
%
The \lstinline[language=Pika]{[Sll]} part of the directive tells the generator which
layouts are used for the arguments. In this case, the function only
has one argument and the \lstinline[language=Pika]{Sll} layout is used. The \lstinline[language=Pika]{Sll} at
the end specifies the layout for the result.

@@

We can generalise our \lstinline[language=Pika]{mapAdd1} to map arbitrary \lstinline[language=Pika]{Int}
functions over a list and then redefine \lstinline[language=Pika]{mapAdd1} using the new
\lstinline[language=Pika]{map}.

@@

This example makes use of \lstinline{instantiate} in two places. In the first case where we have the call
\lstinline[language=Pika]{instantiate [Int] Int f x}, the builtin \lstinline[language=Pika]{Int} layout is used for both the input
and output. In this special case, the \lstinline[language=Pika]{Int} layout shares a name with the \lstinline[language=Pika]{Int} type
that it represents. This is necessary since \lstinline{instantiate} is used for all non-recursive calls that are not constructor applications.

In the second use, \lstinline[language=Pika]{instantiate [Int -> Int, Sll] Sll map add1 xs}, we specify that
the second argument uses the \lstinline[language=Pika]{Sll} layout for the \lstinline[language=Pika]{List} type from \autoref{fig:List-def}. We also give \lstinline[language=Pika]{Sll} as
the layout for the result of the call. Note that it is not necessary to use \lstinline[language=Pika]{instantiate} for the recursive call to \lstinline[language=Pika]{map}. This is because the appropriate layout
is inferred for recursive calls.

It might be surprising that \lstinline[language=Pika]{instantiate} is required in the body of \lstinline[language=Pika]{mapAdd1} since the type signature
of \lstinline[language=Pika]{mapAdd1} suggests that it is layout polymorphic and yet we must pick a specific \lstinline[language=Pika]{List} layout when
we use \lstinline[language=Pika]{instantiate} to call \lstinline[language=Pika]{map}. This is because, in general, a call inside the body of some function \lstinline[language=Pika]{fn} might use
any layout, even layouts that have no relation to the layouts that \lstinline[language=Pika]{fn} is instantiated to.

It is possible to do inference of some layouts, for example in
\lstinline[language=Pika]{mapAdd1} we would usually want to use the same layout as the
argument layout, but we leave this for future work. Another approach
is to introduce type variables that correspond to layouts, as done in
the series of works on the \tname{Dargent}{}
tool~\cite{chen:2023:dargent}.
%
We leave this approach for future work as well.

@@

Two more features that are common in functional languages are \lstinline[language=Pika]{let} bindings and
\lstinline[language=Pika]{if-then-else} expressions. Both of these have straightforward translations into
\SuSLik. A \lstinline[language=Pika]{let} binding corresponds to introducing a new variable with an equality
constraint corresponding to the equality in the \lstinline[language=Pika]{let} binding in the pure part of a \SuSLik{} assertion.

The \lstinline[language=Pika]{if-then-else} construct corresponds to \SuSLik's C-like ternary operator. We can use these features together to implement a \lstinline[language=Pika]{maximum} function:

@@

\noindent
The \lstinline[language=Pika]{let} binding is required here. This is because the recursive predicate
application must use an additional variable, beyond the argument and output
parameters, to store the intermediate result \lstinline[language=Pika]{i} and this is introduced in
\Pika{} using a \lstinline[language=Pika]{let}. It is possible to extend the translator to
automatically introduce these \lstinline[language=Pika]{let} bindings, but we relegate this to future work.

Furthermore, since the \lstinline[language=Pika]{let} is required and the variable introduced by the \lstinline[language=Pika]{let} is used
in the \lstinline[language=Pika]{if} condition, we cannot replace the \lstinline[language=Pika]{if} with a guard. Guards can only occur
at the top-level of a function definition, so it could not occur after a \lstinline[language=Pika]{let}.

@@

A fold is a common kind of operation on a data structure in functional programming, where a binary function
is applied the elements of a data structure to create a summary value. For example, if the binary function
is the addition function, it will give the sum of all the elements of the data structure. The
classic example of such a fold is a fold on a list. In this example, we will write a
right fold over a \lstinline[language=Pika]{List}.

We also demonstrate \lstinline[language=Pika]{Ptr} types and the \lstinline[language=Pika]{addr} builtin operation. This is somewhat similar to an \verb|@| pattern in Haskell,
but it is a low-level construct, indictating how pointers should be used, rather than a high-level construct.
Given a base type $\alpha$ (such as \lstinline[language=Pika]{Int}), \lstinline[language=Pika]{Ptr Int} is a type. This type is
passed around differently in the generated code. In particular, it is passed by reference.
\lstinline[language=Pika]{addr} is the corresponding value-level introduction operation.
\begin{lstlisting}[language=Pika]
%generate fold_List [Int, Sll] (Ptr Int)

@@

\noindent
The compiler produces the following \SuSLik{} specification for \lstinline[language=Pika]{fold_List}:

\begin{lstlisting}[language=SynLang]
predicate fold_List(int i, loc p_x ,loc __r) {
| (p_x == 0) => { __r :-> i }
| (not (p_x == 0)) => {
  p_x :-> x ** (p_x+1) :-> xs ** [p_x,2] **
  func f(p_x, __p_2, __r) **
  fold_List(i, xs, __p_2) **
  temploc __p_2 }
}
\end{lstlisting}
%
In the generated code, \lstinline[language=Pika]{f} is applied to \lstinline[language=Pika]{p_x} rather than
\lstinline[language=Pika]{x} as a result of using \lstinline[language=Pika]{Ptr} and \lstinline[language=Pika]{addr}.

### 2023-10-03 writing/new-paper/related-work.tex
Tidy up to be able to navigate
src is pika source (we ignore c bit)
suslik is the underlying suslik (may need updates)
tests has all the tests we are trying to get to work
writing is the paper we want to submit/revise
various is odd bits and pieces and old C stuff

Furthermore, layout polymorphism also works differently. While
\tname{Dargent} tracks layout instantiations at a type-level with type
variables, in the present work we simply only check to see if a layout
is valid for a given type when type-checking. In particular, we cannot
write type signatures that \textit{require} the same layout in
multiple parts of the type (for instance, in a function type
\verb|List -> List| we have no way at the type-level of requiring that
the argument \verb|List| layout and the result \verb|List| layout are
the same). \Pika{} 2 makes progress towards eliminating this restriction but, as of now, it is layout monomorphic. This more rudimentary approach that \tool currently takes
could be extended in future work.
%
Overall, the examples in the \tname{Dargent} paper tend to focus on
the manipulation of integer values. In contrast, we have focused
largely on data structure manipulation, which follow the primary
motivation of \suslik. \autoref{fig:dargent-comparison} has a comparison of some of the major features of \Pika{} and \tname{Dargent}.

>>>>

Furthermore, layout polymorphism also works differently. While
\tname{Dargent} tracks layout instantiations at a type-level with type
variables, in the present work we simply only check to see if a layout
is valid for a given type when type-checking. In particular, we cannot
write type signatures that \textit{require} the same layout in
multiple parts of the type (for instance, in a function type
\lstinline[language=Pika]{List -> List} we have no way at the type-level of requiring that
the argument \verb|List| layout and the result \verb|List| layout are
the same). \Pika{} 2 makes progress towards eliminating this restriction but, as of now, it is layout monomorphic. This more rudimentary approach that \tool currently takes
could be extended in future work.
%
Overall, the examples in the \tname{Dargent} paper tend to focus on
the manipulation of integer values. In contrast, we have focused
largely on data structure manipulation, which follow the primary
motivation of \suslik. \autoref{fig:dargent-comparison} has a comparison of some of the major features of \Pika{} and \tname{Dargent}.
